---
title: "Species Evaluation Automation Process"
author: "Matthew Van Scoyoc"
date: "`r format(Sys.time(), '%d %B %Y')`"
format:
 html:
     theme: cosmo
     toc: true
     toc-title: Contents
     toc-depth: 2
     toc-location: left
     number-sections: true
     link-external-newwindow: true
 # md: default
 # pdf:
 #     documentclass: report
 #     sansfont: Arial
 #     mainfont: Arial
 #     monofont: Calibri
 #     toc: true
 #     toc-title: Contents
 #     toc-depth: 2
 #     number-sections: true
 #     colorlinks: true
 # docx:
 #     reference-doc: 'style_reference.docx'
 #     mainfont: Arial
 #     monofont: Calibri
 #     toc: true
 #     toc-title: Contents
 #     toc-depth: 2
 #     number-sections: true
bibliography: references.bib
execute:
 echo: true
 message: false
 warning: false
 class-output: 'r'
---

# Introduction

This document details the data acquisition process and the production of automated reports for the Mountain Planning Service Group's (MPSG) Species Evaluations for Species of Conservation Concern (SCC). The MPSG assists National Forests and Grasslands, or planning units, with land management plan revision. Part of the pre-assessment phase of the plan revision process is writing species evaluation documents in preparation for the SCC process. Species evaluations are intended to inform selection of SCC by providing the best available scientific information (BASI) on whether there is "substantial concern about the species’ capability to persist over the long-term in the plan area" (36 CFR 219.9(c)). This process is detailed in draft in the "Species of Conservation Concern Identification Process" document [@usdaforestservice2026]. This document will focus on the data acquisition, or data pull, process and the production of automated species evaluation documents, and is not intended to outline the SCC process.

# Species Evaluation Automation Process

Automation is recommended for this process for two primary reasons: 1) to maintain document content and format consistency, and 2) to reduce workload. Hundreds of Species Evaluations are required for any given planning unit. Automating repeatable steps produces consistent and repeatable documents, and decreases the workload of MPSG staff.

Broadly there are three parts to automating:

1.  Develop a list of species eligible for SCC.\
2.  Gather BASI for each eligible species to assess nativity and occurrence on the planing unit.\
3.  Automate species evaluation documents for species that are native and known to occur on the planning unit.

## Process Preface

### The Use of R Programming Language

MPSG staff uses the R programming language [@rcoreteam2025a] to automate much of this work for repeatability and consistency across planning units. However, automation is not necessarily dependent on R. Many of these processes could be completed by other means. Coding languages, like R, allow MPSG staff to increase efficiency by developing code, or scripted tools, for repeated processes. R code and packages will be referenced in this document to explain the work accomplished.

#### R Packages

R packages are ... <!-- finish paragraph -->

##### The `mpsgSE` Package {#sec-mpsgse}

MPSG staff have been developing the [mpsgSE](https://github.com/fs-scoyoc/mpsgSE) R package [@vanscoyoc2026a] that includes R code functions that conduct steps in this workflow. This section describes key functions in the `mpsgSE` R package.

###### Resolving Taxonomies {#sec-get-taxonid}

The [`get_taxonomies()`](https://github.com/fs-scoyoc/mpsgSE/blob/main/R/get_taxonomies.R) function takes tabular or spatial data with scientific names and returns the data with a taxon ID and full taxonomy added to the origial data provided to the function. It takes one parameter that identifies the variable with the scientific name.

###### Pulling NatureServe State Data {#sec-ns-state-data}

The [`get_ns_state_list()`](https://github.com/fs-scoyoc/mpsgSE/blob/main/R/get_ns_state_list.R) function uses the [natserve](https://github.com/cran/natserv) R package [@chamberlain2026] to pull data from the NatureServe API for a given state. It takes one parameter, the two-character state short code (e.g., "CO" for Colorado) and returns the full list of tracked NatureServe species for that state. This function then uses `get_taxonomies()` to get taxon IDs for each species.

###### Reading Spatial Data into R {#sec-unit-spatail-data}

The [`get_basemap_data()`](https://github.com/fs-scoyoc/mpsgSE/blob/main/R/get_basemap_data.R) function pulls base map data for North and South America, the lower 48 US states, and a given planning unit. Continental and national scale data are acquired using the [`rnaturalearth`](https://github.com/ropensci/rnaturalearth) package [@massicotte2026b]. Forest Service boundary data are acquired from Forest Service ArcGIS Rest Services (<https://apps.fs.usda.gov/arcx/rest/services/EDW>) using [`arcgislayers`](https://r.esri.com/arcgislayers/) package [@parry2025a]. Open Street Map roads data are acquired using the [`osmdata`](https://github.com/ropensci/osmdata) package [@padgham2017].

###### USFWS Birds of Conservation Concern List {#sec-usfws-bcc-list}

This function returns a data frame of bird species that occur in the Birds of Conservation Regions a National Forest or Grassland overlaps.

##### The `rgbif` Package {#sec-rgbif}

The `rgbif` package [@chamberlain2026b; @chamberlain2017b]... <!-- finish paragraph -->

### Taxonomy {#sec-taxonomy}

This workflow would not be possible without addressing taxonomic classification of species names. The classification of living organisms changes frequently as our understanding of the relationships between the boundaries of distinct taxa evolves. This workflow relies on dozens of resources for information that include verbatim scientific names of varying vintage. Many of those resources have their own methods for resolving taxonomy. A single scientific name in one system might refer to completely different sets of organisms in the another system or a group of organisms may have different scientific names applied in a yet another system. This is problematic for automating species evaluations because this process relies heavily on reliably cross-referencing information by species. For example, we use SEINet, Global Biodiversity Information System (GBIF), and state natural heritage programs as sources of species occurrence records. It could be possible that for the grass blue grama (*Bouteloua gracilis*), one source uses the scientific name *Bouteloua gracilis* and another uses a recognized synonym *Chondrosum gracile*. We need a way to recognize both names as the same species. In other words, to make use of both sources, we needed a way to link the same species referred to by different names. It is important that this is done using an algorithm because some data sets return hundreds of thousands of records on a planning unit and manual resolution of taxonomic classification is not possible with large dataset like these.

A variety of resources were evaluated as potential solutions for taxonomic classification. Ultimately, we decided to use the [taxize](https://github.com/ropensci/taxize) R package [@chamberlain2020] because it accesses the GBIF Backbone Taxonomy [@gbifsecretariat] and is easy to use. NatureServe, ITIS, and USDA Plants were candidates for taxonomic classification systems. The ITIS application programming interface (API) was routinely not availble and would fail when trying to resolve long lists of scientific names. NatureServe's API would frequently returned incorrect scientific names and required user input for resolution of each species. USDA Plants was also considered, but is not usable in an automated systems because USDA Plants does not have an API.

The GBIF Backbone Taxonomy [@gbifsecretariata] was chosen because it is a comprehensive taxonomic list with over four million accepted names and three million synonyms that offers flexible and repeatable tools to programmatically validate taxonomies. The GBIF Backbone Taxonomy is based on 105 taxonomic checklists including but not limited to the Catalogue of Life, Integrated Taxonomic Information System, and the International Barcode of Life project. The GBIF Backbone Taxonomy reliably returned the correct scientific name without user input and could reliably be queried in a timely manner.

The process for resolving taxonomy is codified in a function called [`get_taxonomies()`](https://github.com/fs-scoyoc/mpsgSE/blob/main/R/get_taxonomies.R) in the `mpsgSE` R package (see [The mpsgSE Package, Resolving Taxonomies](@sec-get-taxonid)). Often when a scientific name is searched in the GBIF Backbone Taxonomy, the name that is returned might not match the name provided by NatureServe. The taxonomic resolution process developed by MPSG staff is agnostic to scientific names. It instead relies on the a taxonomic identification number, or taxon ID, that is tied to accepted scientific names and associated known synonyms. The taxon ID is used to join the data when combining data sets, and not the scientific names. This process allows us to efficiently use data sets that use different scientific names for the same species.

While the system works well, it is not perfect. Occasionally, scientific names recognized by NatureServe will resolve to the same species in the GBIF Backbone Taxonomy. In this case species are identified with duplicated taxon IDs. Additionally, some species names are not recognized in the GBIF Backbone Taxonomy at all and return invalid taxon IDs. In these cases, manual resolution of scientific name taxonomy will necessary. The `mpsgSE` package maintains a list of species with known taxonomic issues and holds a function to correct known taxonomic problems called

## Building the Eligible Species List

<!-- use pipeline to guide narrative -->

### Planning Unit Spatial Data

Generally, spatial data for the planning unit are read into R from the Forest Service ArcGIS REST Services (<https://apps.fs.usda.gov/arcx/rest/services/EDW>, see @sec-unit-spatail-data). Administrative boundary data are acquired from [Forest Systems Boundaries](https://apps.fs.usda.gov/arcx/rest/services/EDW/EDW_ForestSystemBoundaries_01/MapServer) layer. Forest or grassland administrative boundaries are filtered using the Forest Service Region code and Forest Number code. [Basic ownership](https://apps.fs.usda.gov/arcx/rest/services/EDW/EDW_BasicOwnership_01/MapServer) data are acquired and filtered using Forest Service unit name. The basic ownership data are then filtered by Forest Service-owned lands to delineate the planning area. Three-mile buffers around the administrative boundary is created using the `st_buffer()` function from the [sf](https://r-spatial.github.io/sf/index.html) R package [@pebesma2018; @pebesma2023] to delineate the area of analysis. One-kilometer buffers are created around the planning area boundary using the `st_buffer()` function to capture occurrence records immediately adjacent to the plan area.

In some cases, spatial data for administrative boundaries, basic ownership, plan areas, and buffers are created using ArcGIS Pro [@esriinc.2025]. The administrative boundaries and basic ownership layers are read into ArcGIS Pro from the Forest Service Enterprise Data Warehouse, the corporate spatial data server for the Forest Service. The data are filtered using a [Definition Query](https://pro.arcgis.com/en/pro-app/latest/help/mapping/layer-properties/definition-query.htm) on the "FORESTNAME" field using the name of planning unit to delineate the administrative boundary and basic ownership for the planning unit respectively. The planning area is then delineated using a Definition Query for Forest Service-owned lands from the basic ownership data. Lastly the three-mile and one-kilometer buffers are then created using the [Pairwise Buffer](https://pro.arcgis.com/en/pro-app/latest/tool-reference/analysis/pairwise-buffer.htm) tool.

### Conservation Lists

<!-- update/rewrite -->

Chapter 10 - Assessments of the Forest Service Land Management Planning Handbook [@usdaforestserviceForestServiceHandbook2015b] directs the use of NatureServe in species consideration lists. Therefore, scientific names provided by NatureServe are used for species evaluations. The following data are used to construct a master conservation list for the planning unit to help determine species eligibility for SCC:

-   NatureServe data including Global (G-ranks) and State ranks (S-ranks) and US Fish and Wildife Service (USFWS) Endangered Species Act (ESA) Status,
-   Forest Service Regional Forester's Sensitive Species Lists,
-   USFWS Birds of Conservation Concern List,
-   State Wildlife Action Plans (SWAP),
-   State Threatened and Endangered Species Lists,
-   Bureau of Land Management (BLM) State Sensitive Species Lists, and
-   Neighboring Forest Service Unit SCC Lists.

#### NatureServe State List

The development of the eligible species list begins with pulling state species lists from NatureServe. These data include, species scientific and common names, global and state conservation ranks and USFWS ESA status. To do this, we begin by pulling all species for the state where a unit resides (see [The mpsgSE Package, Pulling NatureServe State Data](@sec-ns-state-data)). In most cases one state species list is needed. But for units that overlap multiple states, each state list must be pulled and then the lists are merged so that unique taxa and species names are preserved.

A data pull for any given state does note determine species eligibility for SCC. Instead it is the total number of species tracked by NatureServe for a given state. To be considered eligible for SCC, a species on a state list must also be found to occupy a given unit (see [Determine Native and Known to Occur](@sec-determine-nko)) and meet the qualifying criteria listed below and outlined in Chapter 10 of the Forest Service Land Management Handbook \[@usdaforestserviceForestServiceHandbook2015\].

Data used from the NatureServe state data pull include:

1.  Scientific and common names,
2.  NatureServe Global Rank (G/T Rank),
3.  NatureServe State Rank (S/T Rank) for a given state, and
4.  USFWS ESA Status (Endangered, Threatened, Candidate, or Under Review).

#### Regional Forester's Sensitive Species Lists

```{r}
rfs_path <- file.path('data-raw/data', '2025_National_RFSS_list.xlsx')
rfs_date <- file.info(bcc_path)$mtime |> as.Date() |> format('%d %B, %Y')
if(is.na(rfs_date)) rfs_date <- "12 January 2026"
```

If a species is ranked as sensitive on the Regional Forester's Sensitive Species List in the region where the planning unit resides that species is considered eligible if it is determined to be native and known to occur on the planning unit (see [Determine Native and Known to Occur](@sec-determine-nko)). Regional Forester's Sensitive Species Lists were acquired from Forest Service Ecosystem Management Coordination staff for every Forest Service Region on `r rfs_date` (personal communication).

#### USFWS Birds of Conservation Concern

```{r}
bcc_path <- file.path('data-raw/data', 'BirdsOfConservationConcern_2024.xlsx')
bcc_date <- file.info(bcc_path)$mtime |> as.Date() |> format('%d %B, %Y')
if(is.na(bcc_date)) bcc_date <- "16 December 2025"
```

A bird species is considered eligible for SCC if that species is listed as breeding on the on the USFWS Birds of Conservation Concern List for Bird Conservation Region where a planning unit resides (see [The mpsgSE Package, USFWS Birds of Conservation Concern List](@sec-usfws-bcc-list)). These data were acquired from the USFWS Birds of Conservation Concern website (<https://www.fws.gov/media/usfws-bird-species-concern>) on `r bcc_date`.

#### State Wildlife Action Plans and Other State Status

States have a variety of wildlife and plant conservation lists for imperiled species. These lists are typically developed by State Natural Resource and Natural Heritage organizations. These conservation lists most commonly come in the form of State Wildlife Actions Plans (SWAP) and state threatened, endangered, and sensitive species lists. The MPSG considers species ranked as Tier 1, Species of Greatest Conservation Need (SGCN), or ranked threatened and endangered eligible for SCC. Some states ranking systems do not adhere to these criteria and may need to be evaluated on a case by case bases. South Dakota, for example, has a three number ranking system where species were given ranks of 1, 2, or 3 and determined that those species ranked 1 would qualify as eligible for SCC. These data are acquired form state natural resource websites or personal communication with state natural resource or natural heritage program staff at the beginning of the SCC process.

#### BLM State Sensitive Species List

The sensitive species list for the state BLM office is acquired if there is BLM land adjacent to the planning unit. Any species that is listed on the state BLM sensitive specieis list will be eligible for SCC if it is determined to be native and known to occur on the planning unit (see [Determine Native and Known to Occur](@sec-determine-nko)). These lists are acquired from the BLM website or direct communication with BLM staff.

#### Neighboring Forest Service Unit SCC Lists

If SCC are developed for a neighboring Forests, they should be considered eligible for SCC if they are native and known on the planning unit. Current SCC lists can be found on the Forest Service planning public site. Some regions make them available grouped by region and others are available at the Forest Level. Below are a few examples:

-   Region 1 - <https://www.fs.usda.gov/detail/r1/landmanagement/planning/?cid=fseprd500402>
-   Region 2, GMUG NF- <https://www.fs.usda.gov/detail/carson/landmanagement/planning/?cid=stelprdb5443166>.
-   Region 4, Manti La Sal NF - <https://www.fs.usda.gov/main/mantilasal/landmanagement/planning>

#### Building the Master Conservation List

A master conservation list is constructed by combining all of the conservation data we have described above using the taxon ID.These data sets are combined using a "full-join" procedure so that species are not dropped from the list during the process. Eligibility in the master conservation list is determined if a species is:

-   G/T 1,2 or 3 ranks,
-   Any S/T 1 or 2 ranks,
-   is on the Regional Foresters Sensitive Species List,
-   is on the USFWS Birds of Conservation Concern list,
-   State SWAP Tier 1 or SGCN ranks,
-   State Threatened or Endangered status,
-   if applicable, is on the State BLM Sensitive Species List, and
-   if applicable, is on the neighboring Forest Service Unit's SCC lists.

#### A Note About Tribal Species of Concern

The MPSG is still developing procedures to incorporate species recognized by "federally recognized Tribes". Those species will be included in the future.

### Occurrence Data

As stated in the 2012 Planning Rule species must be native and known to occur in the planning area. To determine if species are known to occur, open source occurrence databases are queried for species occurrences records on the planning unit. While occurrence records provide the first line of evidence for known to occur we also rely on people who work on the planning unit for additional information. The following datasets are assigned a taxonomic ID using the [`mpsgSE::get_taxonomies()`](https://github.com/fs-scoyoc/mpsgSE/blob/main/R/get_taxonomies.R) function and summarized to the species, subspecies, or variety taxa levels.

#### Global Biodiversity Information Facility

Global Biodiversity Information Facility (GBIF; Global Biodiversity Information Facility 2022a) is a repository of externally sourced species occurrence records from museum collections, academic studies, and citizen science programs. GBIF records requests are staged on GBIF servers and have to be downloaded (see [Getting Occurrence Data From GBIF](https://docs.ropensci.org/rgbif/articles/getting_occurrence_data.html) for details). An R script submits a records request using the 1-km buffer around the planning area boundary to spatially query GBIF records. The data are downloaded in [Darwin Core Archive format](https://www.gbif.org/darwin-core) (GBIF, 2022b) for full data provenance. The data are unzipped and read in to R once the request is available using the `rgbif` package (version 3.7.8, Chamberlain et al. 2023).

#### SEINet

[SEINet](https://swbiodiversity.org/seinet/index.php) is a data portal that provides a suite of data access tools, including species occurrence data from museums, collections, and state and federal agencies. SEINet data are available through an online data portal and are downloaded manually. A polygon box, or well-known text (WKT) footprint, is drawn around the administrative boundary to query species observations using the Taxonomic Criteria search page (<https://swbiodiversity.org/seinet/collections/harvestparams.php>). The query results are downloaded in Darwin Core Archive format, manually unzipped, and a script reads them into R.

#### Integrated Monitoring in Bird Conservation Regions

The [Integrated Monitoring in Bird Conservation Regions (IMBCR)](https://www.birdconservancy.org/what-we-do/science/monitoring/imbcr-program/) is a long-term avian monitoring program coordinated by the [Birds Conservancy of the Rockies](https://www.birdconservancy.org/), and maintains monitoring plots on public lands throughout Forest Service Regions 1-4. IMBCR data spanning 2008-2023 were obtained for Forest Service lands on 12 December, 2023, for use in these analyses. These data were received in and Excel file and an script reads them into R.

#### EO State Data (NHP Data)

State Natural Heritage Programs (NHPs) provide species occurrence data, and habitat and distribution models for federal, state, and non-governmental agencies throughout their state. Element occurrence (EO) spatial data are requested from an HNP which are often provided in a geodatabase. The EO data are read to R using an R script.

#### Forest Service Data

Wildlife, aquatic biota, rare plant, and invasive plant data from the Forest EDW. These data were clipped to the 1-km buffer of the plan area using ArcGIS Pro (version 3.3.1; Environmental Systems Research Institute 2024) and a script reads them into R.

#### A Note on Limitations of Species Occurrence Data

```{=html}
<!-- from data acquisition report

### Occurrence Data {#sec-occurrence_data}

These are occurrence, or presence-only, data and can document the presence of species in an area of interest. Probable or possible presence is not quantified, and most importantly, these data cannot determine the absence or abundance of a species. Lack of species observations is not evidence of absence: a lack of observation may be because the species is rare in that area, or because no one ever looked for it there. Similarly, the number of occurrence records is not an indicator of abundance. For example, showy flowering plants or rare birds will trigger many more occurrence records relative to their abundance than “less interesting” or common species. Lastly, the source external species occurrence datasets often lack information on collection level of effort and on species absences which is essential for estimating abundance.

### Spatial Accuracy {#sec-spatial_accuracy}

Species occurrence locations are not validated and may include coordinate data of varying and undocumented precision or accuracy. The occurrence records from this data pull have geographic location coordinates in longitude and latitude and may include country and state codes and location descriptors. Each observation may include an estimate of the location uncertainty, which may range from a few meters to several kilometers or may be missing. Any use of the occurrence locations should consider the location uncertainty and/or precision noted for each observation. Recent observations may be generated using GPS technology, while coordinates in older observations may have been read from a map, and very old observations usually include only verbal descriptions of the observation location. Some of the verbal descriptions may be “geo-referenced” using current maps and historic location names. Also, some source datasets may have the location coordinates deliberately fuzzed to protect the location of sensitive species.

### Duplicate Records {#sec-duplicate_records}

Some museums and natural history institutions share their data with state heritage programs, GBIF, and SEINet; therefore, we can assume that some observations are duplicated. These analyses do not attempt to reduce duplicated records. Instead, these analyses assemble a comprehensive list of species known to occur on the forest or grassland. It is not appropriate to use these data to estimate populations (see section 2.6.1).

-->
```

#### Build Summary Occurrence Dataset

Once all occurrence records are retrieved, each dataset is summarized by taxon ID with the following fields:

-   "nObs": Total number of observations on Forest Service land.
-   "minYear": Minimum year of observations on Forest Service land.
-   "maxYear": Maximum year of observations on Forest Service land.
-   "occID": A list of occurrence record identification codes if the total number of observations is less than or equal to six.

### Build Preliminary Eligible SCC List

3.  Join all occurrence lists to all potential species eligible filtered list
4.  Filter by those species that have one occurrence record in the planning unit.

The result is a list of species with at least one occurrence record inside the planning area the meats at least one qualifying criteria for consideration as SCC. This list is preliminary because we recognize that his process does not perfectly capture all species that should be considered for SCC and that user input and manual vetting of the list could lead to the addition or subtraction of species.

### Evaluate Preliminary Eligible SCC List

<!-- I think we should add into this process the Units local concern species -->

<!-- One thing I'm unclear on is if this step leads to shorter evaluations or no evaluation at all. -->

#### Determine Native and Known to Occur {#sec-determine-nko .paragraph}

#### Transient Bird List

Per the 2012 planning rule only species that are native to a planning area and not transients or accidental are required to be considered for species evaluation for SCC consideration. As such species must be checked to determine if they are indeed native and not occurring accidentally or due to migration. For many groups of species we can assume that if they are detected and if they have a state ranking or some other ranking that they are likely not accidental nor transients on the plan area. However for migratory birds, some verification should be performed to remove birds that are accidental or transient, including those with detection during migration.

Birds must overwinter or breed within a planning unit to not be considered accidental or transient. To verify that species that are on the initial species list winter or breed on the planning unit we use species distribution modeling provided by the Cornell Lab of Ornothology, through analysis of ebird data [@finkEBirdStatusTrends2023]. Species distribution models are provided through the [ebirdst](https://doi.org/10.2173/ebirdst.2022) R package. Several datasets are available, for our analysis we relied on the 27k resolution "Smooth Range" ("range_smooth") dataset. The dataset provides a variety of polygon spatial layers for each species queried that contain a season attribute that can be used to determine if the species should occur in the planning unit during: breeding, non breeding, pre breeding migration, and post breeding migration or if it occurs year round a single layer representing the resident range. The layers useful to us are those that are breeding, non breeding and resident.

To run the check, each of the native layers (breeding, non breeding or resident) are clipped to the unit boundary. Each range is returned with a TRUE for the range type for that species. So for example if a American goshawks breeding range overlaps overlaps the planning unit goshawk is returned with a true. If any of the non-transient ranges overlaps the planning unit then the variable should_remain_eligible is returned as TRUE. If non of the non-transient ranges overlaps the planning unit then the variable is returned false. These automated checks can be manually verified by an individual if necessary.

This product is stored and read into to inform the final eligible list. Those species with non-transient ranges are kept while those with no resident ranges are removed.

#### Filter those species that need native and known checks

For species with with occurrence records on the unit that could be eligible it must be verified that they are native and known to occur on the planning unit. For many species, there are dozens of records on the planning unit. In these cases we assume that they are native an known to occur on the planning unit (unless they are a bird species that migrates over the area, for instance, and is removed in the previous step). For species with very few occurrence records or with records that are very old we manually check to verify that the occurrence records completely occur on the unit or that that least one record is less than 40 years old.

These checks are run manually. To provide specialists a list of potential checks that need to be conducted we output a spreadsheet with species that have less than 6 occurrence records on the unit or were recorded more than 40 years ago.

Native and known determinations are made and recorded for each species with the following categories with the options:

-   Is the Species Native and Known to Occur?
-   Yes
-   No, the species was introduced or is adventive to the plan area
-   No, there is insufficient taxonomic certainty to identify the observation data to a species
-   No, there is insufficient temporal certainty that the species still occupies the plan area
-   No, there is insufficient spatial certainty that the observation occurred within the plan area
-   No, the species is accidental or transient to the plan area
-   No, the observations represent an expanding species range or irruptive population that is not established in the plan area
-   No, the species is not in the plan area but is in the admin unit therfore is not included in the species overview spatial layer

The checker also writes notes on their determination under "*What is the rationale and supporting BASI for recommending that an observation does not meet the requirements of native to, and known to occur in the plan area?*". Species that are not not Native an Known will remain on the list be will get short species evaluations that only include taxonomic information, eligibility ranking information and a section noting that they are not native and known and why.

#### Species with uncertain taxonomic determinations

Multiple species may be taxonomically identified as the same species with `get_taxonomies()`. Because we cannot determine the taxonomic status programatically, we cannot join them to occurrence lists or status lists. Therefore they must be manually vetted. Additionally, for those species with ambiguous taxonomy we cannot produce automated evaluations.

An example of a species or group of species with ambiguous taxonomic status is the eastern woodrat (*Neotoma floridana*). Natureserve recognizes three subspecies of eastern woodrat: *Neotoma floridana attwateri*, *Neotoma floridana baileyi*, and *Neotoma floridana campestris*. Neither GBIF nor ITIS recognize any of these subspecies but they both recognize the speices *Neotoma floridana*. Because we use the scientific name, varieity or subspecies that NatureServe recognizes, we must recognize all three of these subspecies. However, if we were to link other qualifying lists and occurrence records with the taxon_id provided by GBIF we would use the same ID for occurrence records of all three subspecies and the species. Additionally, if we have no automated way to link the species or subspecies together there is no way to build an automated report as all steps are automated by referencing the `taxon_id`.

#### Add in Species with Local Concern

Unit specialists may have species that they are locally concerned about. Species with local concern are eligible for SCC evaluation if they are native and known to occur on the unit. These species should be added to the eligible list prior to running any further steps. As with every other species species taxonomies should be acquired using the get_taxonomies() function.

One unfortunate part of this step is that it requires that some steps above be rerun. Species do not need any of the rankings above to be considered for this step but all steps should be run to get the status of each local concern species. These steps can rely on the same functionality as those above, but they should be appended prior to running any automated reporting steps.

#### Use user feedback to refine list

Based on manual checks remove species that are:

-   Determined to not be native and known to occur within the unit.
-   Determined to be accidental or transients on the unit

Add species that were identified as having local concern by the unit.

## Retrieve External Data for Automating Reports

### Get Synonyms

For each species we retrieve all recognized synonyms from GBIF for any given taxon_id. We use the function `mpsgSO::get_synonyms()` to query the GBIF Taxonomic Backbone for synonyms. The function takes a vector of taxon_ids and returns a table with all synonmyms associated with each taxon_id. The function relies on the function `rbgif::name_usage(..., data="synonyms")`. This dataset is a stand alone table.

### Get Data for Making Automated Maps

#### Subset Occurrence Spatial Data to Include on Unit and Eligible Species

Occurrence datasets are typically quite large. So that they can more efficiently be loaded into automated reports it is helpful to take the eligible list and subset it to just those species that are on the eligible list and those records that occur on the unit. This step simply filters any given list to those species on the eligible list and then clips the occurrence records to the unit. At the moment it returns a list of two objects, one with all eligible and one with all eligible on the unit identified as `eligible` and `eligible_unit`.

#### Load Species Distribution Models

Species distribution models are needed to understand the context of a species range to any given planning units location. We use a variety of sources of species distribution models to get this information.

##### IUCN

The IUCN provides a variety of species distribution maps for a variety of groups. These maps must be downloaded as shapefiles and then queried for species by taxonomic group (i.e Mammals, Amphibians, Fish, etc). At the moment we use IUCN for all Taxa groups except Birds and Plants. Of all the data sources used for species distribution maps, this is the least efficient source. At the moment, the steps for using these data are:

-   download the data for each taxa group from the IUCN data website <https://www.iucnredlist.org/resources/spatial-data-download>
-   then get the taxonomies off all species in the shapefile with `mpsgSE::get_taxonomies()`.
-   filter the shapefile converted to a table by the species on the eligible lists comparing the taxon_id from each source. We complete this step because filtering a shapefile takes a lot longer (we may want to revise this step).
-   return a list of species with available maps from that data source.
-   if the source has at least one species on the eligible list we then subset the shapefile by those species.
-   For each taxa group we return a shapefile and then those shapefiles are combined.
-   BIEN Plant Maps
-   Ebird Maps

#### Get Other Base Map Information

-   State Boundaries
-   United States - lower 48
-   North America
-   The Americas (North and South America)
-   Get Open Street Map Highways ("highway")
-   motorway
-   trunk
-   primary
-   Clean Unit Names

#### Get NatureServe Habitats From the NatureServe API and Manually Crosswalk to Ecology

#### Retrieve, Clean, and Build Narratives for IMBCR Trend Information

#### Retrieve, Clean, and Build Narratives for Breeding Bird Survey

## Automate Reports

<!-- Look at the qmd and write out each section. (Maybe should wait until we finish revamping the reports) -->

# References

::: {#refs}
:::
